Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-05 article/article.Rmd
Commit 2024-09-05 12:14

In this study our main focus was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the clear-sky irradiance at the time of each GHI
measurement. This depends on the selection of the appropriate atmospheric
parameterization for the RTM simulations. The implementation of the long-term change
in AOD, discussed in section\nobreakspace{}\ref{rtmcs}, allows capturing a large part
of the natural variability of clear-sky GHI. However, the short-term variability of
AOD cannot be taken adequately into account when using monthly values in the model
simulations. We tried different approaches to strengthen the robustness of
the methodology and to compensate for the limited accuracy of the RTM input data and
the unpredictable natural variability of the atmosphere.

@@

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold $E$ is given by:
\begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\text{SZA}$ the solar zenith angle, $E_{\odot}$ is the solar constant at the
mean Sun - Earth distance ($`r round(solar_constant)`\,\text{W}/\text{m}^2$), $r$ is
the actual Sun - Earth distance and $r_\text{m}$ is the mean Sun -- Earth distance of
$1.496\times10^8\,\text{km}$.

>>>>

One of the goals of this study was to quantify the OI related to CEs. A key issue for
achieving this goal is to define a threshold for the CE identification,
representative of the clear-sky irradiance at the time of each GHI measurement. This
depends on the selection of the appropriate atmospheric parameterization for the RTM
simulations. The implementation of the long-term change in AOD, discussed in
section\nobreakspace{}\ref{rtmcs}, allows capturing a large part of the natural
variability of clear-sky GHI. However, the short-term variability of AOD cannot be
taken adequately into account when using monthly values in the model simulations. We
tried different approaches to strengthen the robustness of the methodology and to
compensate for the limited accuracy of the RTM input data and the unpredictable
natural variability of the atmosphere.

@@

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI on horizontal plane at TOA. In this case the threshold $E$ is given by:
\begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\text{SZA}$ the solar zenith angle, $E_{\odot}$ is the solar constant at the
mean Sun - Earth distance ($`r round(solar_constant)`\,\text{W}/\text{m}^2$), $r$ is
the actual Sun - Earth distance and $r_\text{m}$ is the mean Sun -- Earth distance of
$1.496\times10^8\,\text{km}$.

### 2024-09-05 article/article.tex
Commit 2024-09-04 10:14
Commit 2024-09-05 12:14

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1994 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(+46\pm 22\,\text{cases}/\text{year}\), and a
trend in the CE events irradiation of
\(+116.3\pm 68.2\,\text{kJ}/\text{year}\). The peak of the CE events was
observed during May and June. An analysis of the total duration of CE
events, showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour and up to 140 minutes. Finally, we have
detected enhancements above the total solar irradiance at the top of the
atmosphere for the same solar zenith angle of up to
\(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
\(200\,\text{W}/\text{m}^{2}\). The most active season of these extreme
events is spring until early summer and autumn.

@@

According to \citet{Kazadzis2007}, in the period 1997 -- 2005 the mean
AOD at \(340\,\text{nm}\) is \(0.403\) with a change of
\(-3.8\pm0.93\,\%\) per year, corresponding to a change of \(0.0153\)
per year. Using an Ångström coefficient \(\alpha = 1.6\), this
translates to a change in the Ångström coefficient \(\beta\) of
\(0.00272\) per year (or \(\beta=0.084\) in 1997 and \(\beta=0.059\) in
2005). Simulations with uvspec for the above Ångström coefficients,
assuming constant WC of \(15.6\,\text{mm}\) taken from the Cimel, and
for a SZA of \(55^\circ\) reveal a change of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean clear-sky
irradiance, and finally the change of \(+0.14\,\%\) per year. We applied
these two long-term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the clear-sky irradiance for the whole period of study. The rate of
changes in the two subperiods are different, but reasonable, since the
effect of the enforced air pollution abatement measures in the area in
the 1990s had greater effects in the first years of application.

@@

A sub-category of the CE events that is often discussed in the
literature \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI at the top of the
atmosphere. In this case the threshold \(E\) is given by:
\begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation} where: \(\text{SZA}\) the solar zenith angle,
\(E_{\odot}\) is the solar constant at the mean Sun - Earth distance
(\(1361\,\text{W}/\text{m}^2\)), \(r\) is the actual Sun - Earth
distance and \(r_\text{m}\) is the mean Sun -- Earth distance of
\(1.496\times10^8\,\text{km}\).

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(1.78\,\%\) were identified as CE events and \(0.04\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(345.9\,\text{W}/\text{m}^2\) or
\(32.3\,\%\) above the threshold. The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(286.1\,\text{W}/\text{m}^2\). Both cases are ECE events with
\(136.8\,\text{W}/\text{m}^2\) and \(164.5\,\text{W}/\text{m}^2\) above
the TSI at TOA for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+46\pm 22\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.3 \pm 68.2\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0.03\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OI for the entire period is
\(+40.1\pm 2.5\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(40.1\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 28182 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013 in a SZA range of \(52.1^\circ\) --
\(77.9^\circ\). Figure\nobreakspace{}\ref{fig:ceg-duration-distribution}
shows the frequency distribution of the CE events according to their
duration. We conclude that although some groups of events last for more
than an hour, about \(80\,\%\) have duration of less than 5 minutes.

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+45.6\pm 22\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+116.3\pm 68.2\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.

>>>>

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1994 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(+85\pm 20\,\text{cases}/\text{year}\), and a
trend in the CE events irradiation of
\(+226.4\pm 62.6\,\text{kJ}/\text{year}\). The peak of the CE events was
observed during May and June. An analysis of the total duration of CE
events, showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour and up to 140 minutes. Finally, we have
detected enhancements above the total solar irradiance at the top of the
atmosphere for the same solar zenith angle of up to
\(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
\(200\,\text{W}/\text{m}^{2}\). The most active season of these extreme
events is spring until early summer and autumn.

@@

According to \citet{Kazadzis2007}, in the period 1997 -- 2005 the mean
AOD at \(340\,\text{nm}\) is \(0.4\) with a change of
\(-3.8\pm0.93\,\%\) per year, corresponding to a change of \(0.0153\)
per year. Using an Ångström coefficient \(\alpha = 1.6\), this
translates to a change in the Ångström coefficient \(\beta\) of
\(0.00272\) per year (or \(\beta=0.084\) in 1997 and \(\beta=0.059\) in
2005). Simulations with uvspec for the above Ångström coefficients,
assuming constant WC of \(15.6\,\text{mm}\) taken from the Cimel, and
for a SZA of \(55^\circ\) reveal a change of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean clear-sky
irradiance, and finally the change of \(+0.14\,\%\) per year. We applied
these two long-term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the clear-sky irradiance for the whole period of study. The rate of
changes in the two subperiods are different, but reasonable, since the
effect of the enforced air pollution abatement measures in the area in
the 1990s had greater effects in the first years of application.

@@

A sub-category of the CE events that is often discussed in the
literature \citep{Cordero2023, Martins2022, Yordanov2015}, are the
extreme cloud enhancement (ECE) events. These are cases of CE where the
measured intensity of the irradiance exceeds the TSI on horizontal plane
at TOA. In this case the threshold \(E\) is given by: \begin{equation}
\text{ECE}: E > \cos(\text{SZA}) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation} where: \(\text{SZA}\) the solar zenith angle,
\(E_{\odot}\) is the solar constant at the mean Sun - Earth distance
(\(1361\,\text{W}/\text{m}^2\)), \(r\) is the actual Sun - Earth
distance and \(r_\text{m}\) is the mean Sun -- Earth distance of
\(1.496\times10^8\,\text{km}\).

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(1.68\,\%\) were identified as CE events and \(0.04\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(338.4\,\text{W}/\text{m}^2\) or
\(31.4\,\%\) above the threshold. The strongest CE event of \(49.9\,\%\)
above the clear sky threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(286.9\,\text{W}/\text{m}^2\). Both cases are ECE events with
\(136.8\,\text{W}/\text{m}^2\) and \(164.5\,\text{W}/\text{m}^2\) above
the TSI at TOA for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+85\pm 20\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+226.4 \pm 62.6\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0.12\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OI for the entire period is
\(+39.7\pm 2.8\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(39.7\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 26837 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013 in a SZA range of \(52.1^\circ\) --
\(77.9^\circ\). Figure\nobreakspace{}\ref{fig:ceg-duration-distribution}
shows the frequency distribution of the CE events according to their
duration. We conclude that although some groups of events last for more
than an hour, about \(80\,\%\) have duration of less than 5 minutes.

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+85.3\pm 20\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+226.4\pm 62.6\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.
